---
title: "rf_to_network"
format: html
---

# Setup

Forests from various packages may have somewhat different formats. I'll try XGBoost maybe.

```{r}
pacman::p_load(tidyverse, randomForest)

#load data
Auto <- ISLR2::Auto
glimpse(Auto)

forest1 <- randomForest(formula = mpg ~ . -name, data = Auto, importance = TRUE, ntree = 3)# make a small forest 
forest2 <- randomForest(formula = as.factor(mpg) ~ . -name, data = Auto, importance = TRUE)
out <- edges.randomForest(forest1)
```


# Explore structure

```{r}
ff <- forest1$forest
# View(ff)
ff |>
  summary()
ff2 <- forest2$forest
ff$nodestatus[,1]
ff$leftDaughter[,1]
ff$rightDaughter[,1]
ff$nodepred[,1]
ff$bestvar[,1]
ff$xbestsplit[,1]

tree1 <- getTree(forest1, 1)# using this function is probably more future-proof than relying on object staying the same
tree1 |> head()
tree1[,"status"] |> unique()
unique(tree1)
# why are these -3 status? docs claim it should be -1 or 1
tree2 <- getTree(forest2, 1)
tree2 |> head()

# unique()

```

# Convert one tree

```{r}
# efficiency doesn't matter, write it first then make it fast

convert_tree <- function(treenum){
  tree1 <- getTree(forest1, treenum)
  tree1 <- as.data.frame(tree1)
  tree1$index <- c(1:nrow(tree1))
  
  parent_index <- tree1$`left daughter` != 0
  edgelist <- data.frame(source = c(tree1[parent_index,"index"], tree1[parent_index,"index"]),
                         target = c(tree1[parent_index,"left daughter"], tree1[parent_index,"right daughter"]),
                         split_var = c(tree1[parent_index,"split var"], tree1[parent_index,"split var"]),
                         split_point = c(tree1[parent_index,"split point"], tree1[parent_index,"split point"]),
                         prediction = c(tree1[parent_index,"prediction"], tree1[parent_index,"prediction"]))
  # structure is here (for one tree). but u can add the variable names I guess
  # you can do some kind of default thing.
  # idk uhhhhh labels for prediction?????
  # 
  return(edgelist)
}
# edgelist$split_var |> unique()
forest_edge <- data.frame()
# is most efficient way adding to a list? 
for(i in 1:forest1$ntree){
  forest_edge <- rbind(forest_edge, convert_tree(i))
  
}

# do for all trees then attach varnames

```

# Just exploring converted data

(Not part of package I'm just interested in what one could do with it)

note xgboost plot inspired by https://wellecks.wordpress.com/2015/02/21/peering-into-the-black-box-visualizing-lambdamart/ 


```{r}
forest_edge <- out
g <- igraph::graph_from_data_frame(forest_edge)
# graph is wrong bc source target aren't uniquely identified
library(ggraph)
ggraph(g) +
  geom_edge_link(angle_calc = 'along') +
  # geom_point(aes(x = x, y = y)) +
  # geom_node_label(size = 2, repel = TRUE) +
  theme_void()
```


```{r}
library(igraph)
library(tidygraph)
g2 <- graph_from_data_frame(forest_edge)
g2 |>
  as_tbl_graph() |>
  mutate(dist_to_center = node_distance_to(node_is_center())) %>% 
    ggraph(layout = 'kk') + 
    geom_edge_link() + 
    geom_node_point(aes(size = dist_to_center), colour = 'steelblue') + 
    scale_size_continuous(range = c(6, 1)) + 
    theme_graph()
```

# Testing some stuff out

What follows is AI - what can the ai do w this edgelist?

Not important - my contribution is just the edgelist converter.

```{r}
library(tidygraph)
library(ggraph)
library(dplyr)

edgelist <- forest_edge
# Simple version using just source/target columns
visualize_rf_network_simple <- function(edgelist, min_connections = 10) {
  
  # Create basic graph from source/target
  graph <- tbl_graph(
    edges = edgelist %>% select(from = source, to = target),
    directed = TRUE
  )
  
  # Calculate node metrics
  graph <- graph %>%
    activate(nodes) %>%
    mutate(
      degree = centrality_degree(),
      betweenness = centrality_betweenness()
    )
  
  # Sample if too large
  if (ecount(graph) > 5000) {
    graph <- graph %>%
      activate(nodes) %>%
      filter(degree >= min_connections) %>%
      activate(edges) %>%
      filter(!is.na(from) & !is.na(to))
  }
  
  ggraph(graph, layout = 'fr') +
    geom_edge_link(alpha = 0.3, edge_colour = "gray40") +
    geom_node_point(aes(size = degree), color = "steelblue", alpha = 0.7) +
    scale_size_continuous(range = c(1, 8)) +
    theme_graph() +
    labs(title = "Random Forest Network Structure")
}

# Version with split_var grouping
visualize_rf_grouped <- function(edgelist, min_edge_count = 20, top_vars = 30) {
  
  # Group edges by split_var
  var_transitions <- edgelist %>%
    group_by(split_var) %>%
    mutate(
      next_var = lead(split_var)
    ) %>%
    ungroup() %>%
    filter(!is.na(next_var)) %>%
    count(split_var, next_var, name = "weight") %>%
    filter(weight >= min_edge_count)
  
  # Create graph from variable transitions  
  graph <- tbl_graph(
    edges = var_transitions %>% rename(from = split_var, to = next_var),
    directed = TRUE
  )
  
  # Add node statistics
  graph <- graph %>%
    activate(nodes) %>%
    mutate(
      in_deg = centrality_degree(mode = "in"),
      out_deg = centrality_degree(mode = "out"),
      total_deg = centrality_degree(mode = "all")
    ) %>%
    # Keep top variables
    top_n(top_vars, total_deg)
  
  # Keep only edges between remaining nodes
  graph <- graph %>%
    activate(edges) %>%
    filter(!edge_is_incident(which(.N()$name %in% .N()$name[is.na(.N()$total_deg)])))
  
  # Visualize
  ggraph(graph, layout = 'fr') +
    geom_edge_link(
      aes(width = weight, alpha = weight),
      arrow = arrow(length = unit(2, 'mm'), type = "closed"),
      end_cap = circle(2, 'mm'),
      colour = "gray30"
    ) +
    geom_node_point(
      aes(size = total_deg),
      color = "steelblue", 
      alpha = 0.8
    ) +
    geom_node_text(aes(label = name), repel = TRUE, size = 3) +
    scale_edge_width_continuous(range = c(0.5, 3)) +
    scale_edge_alpha_continuous(range = c(0.3, 0.9), guide = "none") +
    scale_size_continuous(range = c(3, 12)) +
    theme_graph() +
    labs(
      title = "Variable Transition Network",
      subtitle = sprintf("Top %d variables, edges with %d+ transitions", 
                        top_vars, min_edge_count)
    )
}

# Version with prediction values
visualize_rf_with_predictions <- function(edgelist, sample_size = 2000) {
  
  # Sample if needed
  if (nrow(edgelist) > sample_size) {
    edgelist <- edgelist %>% sample_n(sample_size)
  }
  
  # Bin predictions for coloring
  edgelist <- edgelist %>%
    mutate(
      pred_group = cut(prediction, 
                       breaks = quantile(prediction, probs = c(0, 0.33, 0.67, 1)),
                       labels = c("Low", "Med", "High"),
                       include.lowest = TRUE)
    )
  
  # Create graph
  graph <- tbl_graph(
    edges = edgelist %>% select(from = source, to = target, pred_group),
    directed = TRUE
  )
  
  # Color edges by prediction group
  ggraph(graph, layout = 'stress') +
    geom_edge_link(
      aes(edge_colour = pred_group),
      alpha = 0.4
    ) +
    geom_node_point(size = 0.5, alpha = 0.5) +
    scale_edge_color_manual(values = c("Low" = "blue", "Med" = "gray", "High" = "red")) +
    theme_graph() +
    labs(title = "Random Forest Flow by Prediction Values")
}

# Usage:
plot1 <- visualize_rf_network_simple(edgelist)
plot2 <- visualize_rf_grouped(edgelist)
plot3 <- visualize_rf_with_predictions(edgelist)
```

```{r}
plot1
plot2
plot3
```
