---
title: "Reformat Tree"
format: html
---



## Put in a different classification tree?

# 1. Classification Trees

## a. Load data
```{r}
data(Auto, package = "ISLR2")
dplyr::glimpse(Auto)
```

## b. Add categorical variable Eco

```{r}
Auto$eco <- as.factor(ifelse(Auto$mpg > mean(Auto$mpg), "Economy", "Consuming"))
dplyr::glimpse(Auto)
```

```{r}
table(Auto$eco)
```

Above are the numbers of observations for Economy and Consuming levels.

## c. Fit tree

```{r}
library(tree)
tree(eco~mpg, data = Auto)
```


```{r}
mean(Auto$mpg)
```

It makes sense that the split is near the mean because the mean of mpg was the defined split point for the eco variable. With this split point, the tree can achieve 100% accuracy trivially.

## d. Fit second tree

```{r}
eco_tree <- tree(eco~. -name -mpg, data = Auto)
eco_tree

summary(eco_tree)
```

The root node has a split based on the proportions of each class in the data. The first split is based on displacement above or below 190.5.

## e. Visualize

```{r}
#| eval: false
# investigating the tree object
as.character(eco_tree)
class(eco_tree)
eco_tree$frame
eco_tree
```

```{r}
#plot could be better looking
plot(eco_tree)
text(eco_tree)
```

## Reformat!!

```{r}
#| eval: false
#| include: false
df <- eco_tree$frame
# I believe this part may be all I need to construct an edge list.
# not about efficiency at this stage - you can use a for loop, whatever


#initialize empty edge list
edges <- data.frame(from = integer(0), to = integer(0))
n <- nrow(df)# i assume this is more efficient than counting the rows every time


# construct edges labelled w rownums
#STILL WRONG LMAO

#reconsider strat
# each node is a child once except the first one
# Each non-leaf node is a parent twice

# keep index
df$index <- c(1:n)

#all parent nodes
noleaf <- subset(df, var != "<leaf>")
parents <- c(noleaf$index, noleaf$index)

#all child nodes 
children <- c(2:n)

for (i in 1:n){
  if (i %in% parents){
    {
      
    }
  }
}

# shit do i need to do recursion or something i was not a cs major
# there's usually a way to do stuff without it right?

add_child <- function(child){
  if (!(child %in% parents)){# if child is not a parent
    
    return()
  } else {
    
  }
}

##
#starting over again lmao why dont i do something for project? idk
for (i in 1:n) {
  if(df$var[i] != "leaf"){
    edges <- rbind(edges, data.frame(from = i, to = i+1))
  } else {
    
  }
}





##

for (i in 1:n) {
  if(df$var[i+1] == "leaf"){
    edges <- rbind(edges, data.frame(from = i, to = i+1))
    edges <- rbind(edges, data.frame(from = i, to = i+2))
  } else {
    if (df$var[i] != "<leaf>") {
      #inefficient
      
    } 
  }
}

# Now to add other information - there are decisions to make here
# default let's just match plot.tree viz - node is labelled with split

edges$label <- NA
for (i in 1:nrow(edges)) {
  parent_node <- edges$from[i]
  child_node <- edges$to[i]
  var <- df$var[parent_node]
  split <- df$splits[parent_node]
  
  edges$label[i] <- ifelse(child_node %% 2 == 0,
                           paste(var, df$splits[parent_node,1]),
                           paste(var, df$splits[parent_node,2]))
  
}


```


```{r}

df <- eco_tree$frame
# I believe this part may be all I need to construct an edge list.
# not about efficiency at this stage - you can use a for loop, whatever
#initialize empty edge list
edges <- data.frame(from = integer(0), to = integer(0))#efficiency?
n <- nrow(df)# i assume this is more efficient than counting the rows every time
# construct edges labelled w rownums
#STILL WRONG LMAO
df$index <- c(1:n)

is_leaf <- df$var == "<leaf>"

parent_stack <- c()
children_count <- c()

for (i in 1:n){
  # If we have a parent waiting for children
    if (length(parent_stack) > 0) {
      parent_row <- parent_stack[length(parent_stack)]
      parent_idx <- df$index[parent_row]
      
      
      edges <- rbind(edges, data.frame(
        from = parent_idx,
        to = i
      ))
      
      # Increment children count for this parent
      children_count[length(children_count)] <- children_count[length(children_count)] + 1
      
      # If parent has both children, pop it from stack
      if (children_count[length(children_count)] == 2) {
        parent_stack <- parent_stack[-length(parent_stack)]
        children_count <- children_count[-length(children_count)]
      }
    }
  
  #add current node to parent stack and indicate 0 children
  if(!is_leaf[i]){
    parent_stack <- c(parent_stack, i)
    children_count <- c(children_count, 0)
  }
}


### NOW add labels
edges$label <- NA
for (i in 1:nrow(edges)) {
  parent_node <- edges$from[i]
  child_node <- edges$to[i]
  var <- df$var[parent_node]
  split <- df$splits[parent_node]
  
  edges$label[i] <- ifelse(child_node %% 2 == 0,
                           paste(var, df$splits[parent_node,1]),
                           paste(var, df$splits[parent_node,2]))
  
}

```



```{r}
# make test directed igraph
library(igraph)
library(tidygraph)
# g <- graph_from_data_frame(edges, directed = TRUE)
# layout <- layout_as_tree(g)


# plot(g, edge.label = E(g)$label, layout = layout)


library(ggraph)
ggraph::ggraph(g) +
  geom_edge_link(aes(label = label), angle_calc = 'along',
                 label_dodge = unit(2.5, 'mm'),
                 label_size = 2,
                 arrow = arrow(length = unit(4, 'mm')), 
                 end_cap = circle(3, 'mm')) +
  geom_point(aes(x = x, y = y)) +
  geom_node_label(size = 2, aes(label = name), repel = TRUE) +
  theme_void()
```