---
title: "xgboost"
format: html
---

# Setup

```{r}
pacman::p_load(xgboost)
dplyr::glimpse(mtcars)
```

```{r}
y <- mtcars$mpg
x <- mtcars[,c("cyl", "disp", "hp", "drat", "wt", "qsec")]
```

```{r}
xg <- xgboost(x, y)
```

This is an external pointer! I guess that makes sense. It's efficient and stuff so it's C/C++?


# Explore structure

Do I convert this first? Is there like an RCPP thing I need to do or will this be simple

also - if they have a graphviz/diagrammer plot function, 

```{r}
raw <- xgb.save.raw(xg)
# no idea what this means
dump <-xgb.dump(xg, dump_format = "json")[[1L]]
dump
```
```{r}
xg_j <- jsonlite::fromJSON(dump)
```

```{r}
xg_dt <- xgb.model.dt.tree(model = xg)
```

JSON seems straightforward!! It's a hierarchical structure so pretty intuitive. Something for doing this might even already exist, but I want my package to support lots of inputs + don't want too many dependencies. On the other hand, dt is more similar to the other ones I've done. WAIT that's literally way easier doing that.

Note to self: remember, you can make more efficient later (eg do in C++ lol).

Is this going to be a headache without those unnest functions? :( Oh nvm it's easy.

I think the next issue is: how do I handle node attributes? It is possible to put that information in an edgelist. But I need to know the most convenient way. I think I may have to mess around w igraph for a while.


```{r}

df <- data.frame(source = c(xg_dt$ID, xg_dt$ID),
                 target = c(xg_dt$Yes, xg_dt$No))
# lol that's literally it some of these are way more convenient
# now I just have to add the other columns.
# also I can format IDs this way for randomForest package
```


