---
title: "Visualizing models as networks"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Visualizing models as networks}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Tree-based models — decision trees, random forests, boosted ensembles —
are networks: nodes connected by edges. **networkformat** converts them
into graph objects so you can draw them with
[ggraph](https://ggraph.data-imaginist.com/) like any other network.

```{r setup, message = FALSE}
library(networkformat)
library(ggraph)
library(tidygraph)
```

# From model to plot

Fit a model, convert it, plot it:

```{r tree-minimal, eval = requireNamespace("tree", quietly = TRUE), fig.width = 9, fig.height = 5.5, out.width = "100%", message = FALSE}
tr <- tree::tree(Species ~ Sepal.Length + Sepal.Width, data = iris)

as_tbl_graph(tr) |>
  ggraph(layout = "tree") +
  geom_edge_diagonal(colour = "grey60") +
  geom_node_label(aes(label = label, fill = is_leaf), size = 3) +
  scale_fill_brewer(palette = "Set2", name = "Terminal node") +
  theme_graph(base_family = "sans")
```

`as_tbl_graph()` returns a tidygraph `tbl_graph`, which is what `ggraph()`
plots. Every node arrives with attributes ready to map — `label` (a
display label the package builds for you), `var` (the split variable),
`yval` (the prediction), `is_leaf`, `n` (observations reaching the node),
and `depth`. Edges carry the parsed split rule: `split_var`, `split_op`,
and `split_point`.

If you prefer igraph, `as.igraph(tr)` returns an igraph object with the
same attributes.

# Under the hood: edgelists and nodelists

`as_tbl_graph()` is a convenience wrapper. The package's core functions
return plain data frames you can inspect, filter, or join before building
a graph:

```{r tree-edgelist, eval = requireNamespace("tree", quietly = TRUE)}
edges <- edgelist(tr)
nodes <- nodelist(tr)
head(edges, 3)
head(nodes, 3)
```

These are ordinary data frames — the node `name` values match the edge
`from`/`to` values, so they slot directly into
`igraph::graph_from_data_frame(edges, vertices = nodes)`. Use this route
when you want to prune edges, compute graph statistics, or attach your
own node attributes first. `vignette("edgelist-nodelist")` covers the
full workflow.

# Polishing the tree plot

The node and edge attributes support a more annotated rendering: color
internal nodes by split variable and leaves by predicted class, and put
the split condition on each edge.

```{r tree-polished, eval = requireNamespace("tree", quietly = TRUE), fig.width = 9, fig.height = 6}
tg <- as_tbl_graph(tr) |>
  mutate(fill_var = ifelse(is_leaf, yval, var))

ggraph(tg, layout = "tree") +
  geom_edge_diagonal(colour = "grey60", width = 0.5) +
  geom_edge_diagonal(
    aes(label = ifelse(is.na(split_point), "", paste0(split_op, split_point))),
    angle_calc = "along", label_dodge = unit(3, "mm"),
    label_size = 2.5, label_colour = "grey30", colour = NA) +
  geom_node_label(
    aes(label = label, fill = fill_var),
    size = 2.5, colour = "white", fontface = "bold",
    label.padding = unit(0.25, "lines"), label.r = unit(0.2, "lines")) +
  scale_fill_brewer(palette = "Dark2", name = NULL) +
  theme_graph(base_family = "sans") +
  theme(legend.position = "bottom") +
  labs(title = "Iris classification tree",
       subtitle = "Internal nodes: split variable. Leaves: predicted class. Edges: split rule.")
```

The same recipe works for `rpart` models — `as_tbl_graph(fit)` produces
the same node and edge attributes, so the plotting code transfers
unchanged.

# Random forests

`as_tbl_graph(rf)` puts every tree in one graph as disconnected
components, with a `treenum` attribute on both nodes and edges.

```{r rf-fit, eval = requireNamespace("randomForest", quietly = TRUE), message = FALSE}
set.seed(42)
rf <- randomForest::randomForest(Species ~ ., data = iris,
                                 ntree = 9, maxnodes = 6)
tg_rf <- as_tbl_graph(rf)
tg_rf
```

## All trees at once

Faceting on `treenum` draws the forest as small multiples — a quick
structural overview of how much the trees differ:

```{r rf-facets, eval = requireNamespace("randomForest", quietly = TRUE), fig.width = 10, fig.height = 8}
tg_rf <- tg_rf |>
  mutate(pred_class = rf$classes[prediction],
         fill_var = ifelse(is_leaf, pred_class, split_var_name))

ggraph(tg_rf, layout = "tree") +
  geom_edge_link(colour = "grey70", width = 0.4) +
  geom_node_point(aes(colour = fill_var), size = 3) +
  scale_colour_brewer(palette = "Dark2", name = NULL) +
  facet_nodes(~ treenum, scales = "free") +
  theme_graph(base_family = "sans") +
  theme(legend.position = "bottom") +
  labs(title = "A random forest as small multiples",
       subtitle = "One panel per tree. Points: split variable (internal) or predicted class (leaf).")
```

At this scale, labels would overlap — points colored by split variable
show each tree's shape and which predictors it uses. For a forest with
hundreds of trees, facet a sample: `filter(treenum %in% sample(rf$ntree, 9))`.

## One tree in detail

Pass `treenum` to pull a single tree, then label it fully:

```{r rf-single, eval = requireNamespace("randomForest", quietly = TRUE), fig.width = 8, fig.height = 5}
tg1 <- as_tbl_graph(rf, treenum = 1) |>
  mutate(pred_class = rf$classes[prediction],
         display_label = ifelse(is_leaf, pred_class,
                                paste0(split_var_name, "\n< ", round(split_point, 1))),
         fill_var = ifelse(is_leaf, pred_class, split_var_name))

ggraph(tg1, layout = "dendrogram") +
  geom_edge_elbow(colour = "grey65", width = 0.4) +
  geom_node_label(
    aes(label = display_label, fill = fill_var),
    colour = "black", size = 2.5,
    label.padding = unit(0.25, "lines"), label.r = unit(0.2, "lines")) +
  scale_fill_brewer(palette = "Set2", name = NULL) +
  theme_graph(base_family = "sans") +
  theme(legend.position = "bottom") +
  labs(title = "Tree 1 from the random forest")
```

# Boosted models

`xgb.Booster` and `gbm` models convert the same way. Their node
attributes differ slightly — xgboost nodes carry `feature`, `quality`,
and `cover`; gbm nodes carry `split_var_name`, `prediction`, and
`error_reduction` — but `label` and `is_leaf` are always present, so the
minimal recipe is identical:

```{r xgb, eval = requireNamespace("xgboost", quietly = TRUE), fig.width = 8, fig.height = 5, message = FALSE}
dtrain <- xgboost::xgb.DMatrix(
  as.matrix(iris[, 1:4]),
  label = as.integer(iris$Species == "virginica"))
xgb <- xgboost::xgb.train(
  params = list(max_depth = 3, objective = "binary:logistic"),
  data = dtrain, nrounds = 2, verbose = 0)

as_tbl_graph(xgb, treenum = 1) |>
  ggraph(layout = "tree") +
  geom_edge_diagonal(colour = "grey60") +
  geom_node_label(aes(label = label, fill = is_leaf), size = 2.5) +
  scale_fill_brewer(palette = "Set2", name = "Terminal node") +
  theme_graph(base_family = "sans") +
  labs(title = "First tree of an xgboost model")
```

# Where next

- `vignette("edgelist-nodelist")` — the full `edgelist()` / `nodelist()`
  workflow for data frames, lists, and vectors, including edge filtering
  and graph statistics.
- `vignette("networkformat")` — package overview and the `courses`
  data-frame example.
- The [complete method reference](https://jessebrandtdata.github.io/networkformat/articles/complete-method-reference.html)
  on the package site shows every supported model type.
